Draft Exploratory Analysis

### 2024-12-02 Data_Cleaning_20241031.Rmd
Add files via upload

---
title: "ECNS 560 Project - Bison Population"
author: "Kyle Froisland"
date: "2024-10-16"
output: html_document
---

@@

# Plot with custom y-axis labels
ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "loess") +
  scale_y_log10(labels = comma) +  # Use comma format for log scale labels
  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
       x = "Year", 
       y = "Count / Cost",
       color = "License Type") +
  theme_minimal()
```

>>>>

<<<<<<< HEAD
---
title: "ECNS 560 Project - Bison Population"
author: "Kyle Froisland"
date: "2024-10-16"
output: html_document
---

#Load States:
states = st_read("C:/Users/kdfro/Documents/MSU/GitHub/course-materials/lecture-slides/12-Spatial/data/us_states_contiguous/states_contiguous.shp")
```
Fix States and Ranges:
```{r}
#Check CRS
st_crs(states)
st_crs(ranges)

graph5_data = merged_bison_data |>
  filter(state_code == "MONTANA" | state_code == "TEXAS" | state_code == "OKLAHOMA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA" | state_code == "NEBRASKA" | state_code == "KANSAS" | state_code == "OKLAHOMA" | state_code == "IDAHO" | state_code == "COLORADO" | state_code == "NEW MEXICO")

#Populations for all states of interest. There seems to be larger decreases in population around western planes states than southwestern states.
ggplot(graph5_data, aes(x = year, y = total_population)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Bison Population by State - All States of Interest", x = "Year", y = "Bison Population") +
  facet_wrap(~state_code)
#This is also biased because southwestern states started with much lower populations to begin with.

Research Question 1: Can we predict percent decrease in population levels based on Latitude of their centroid (Proxied by proximity to Saskatchewan)?
```{r}
#Load Canada Shapefile:
Canada = st_read("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/ca_shp/ca.shp")

#Convert GRS:
Canada2 = st_transform(Canada, crs = st_crs(states2))
stopifnot(st_crs(Canada2) == (st_crs(states2)))

#Find Centroids of states of interest:
state_centroids = st_centroid(states2)

#Plot to confirm:
tm_shape(states2) + tm_borders() +
  tm_shape(states2) + tm_dots(col = "red")

#Make a variable for proximity of centroid to Saskatchewan:
distance_pop_data = state_centroids |>
  mutate(distance = st_distance(state_centroids$geometry, Canada2[3,4]))

#Merge data with population, create variables for distance from Canada and population decline:
distance = as.data.frame(distance_pop_data) |>
  rename(state_code = STATE) |>
  mutate(state_code = toupper(state_code)) |>
  left_join(bison_data_merge, by = "state_code") |>
  group_by(state_code) |>
  mutate(difference = abs(total_population - total_population[year == 2002])/total_population[year == 2002]) |>
  filter(year == 2017)

#Plot Proximity vs Population Decrease:
ggplot(distance, aes(x = log(distance), y = difference)) + 
  xlab("log(distance) from Canada") +
  ylab("Percent Decrease in Bison Population") +
  labs(title = "Latitude and Bison Population Decline") +
  geom_point()
#There does not seem to be as clear of a relationship from this plot as I had originally expected. The predictive power of this variable seems tenuous.
```

Research Question 2: Can we predict population decline using Latitude and hunting data? This probably won't work due to limited data points...
```{r}
library(tidyverse)
library(tidymodels)
library(ISLR)
library(glmnet)

#Merge Distance data set with hunting data:
distance_hunting = left_join(distance, hunting_data_tidy, by = c("state_code" = "state_code", "year" = "year")) |> as.data.frame()

#Create Predictive Model:
#First, split into train and test data
set.seed(123)
bison_split = distance_hunting |> initial_split(prop = 0.7)
bison_train = distance_hunting |> training()
bison_test = distance_hunting |> testing()

#Define a Recipe:
bison_recipe = recipe(difference ~ disatnce + total_population + Paid Hunting License Holders, data = bison_train) |> 
  update_role(GEOID, new_role = "id") |> 
  step_poly(all_numeric_predictors(), degree = 2) |> #create second term for each numeric var that is a quadratic
  step_interact(~all_numeric_predictors():all_nominal_predictors()) |> #create interactions between numeric and indicator variables
  step_dummy(all_nominal_predictors()) |> #create indicators for each categorical or character variable
  step_nzv(all_predictors())#remove predictors with nearly 0 variance... otherwise not very reliable because difficult to pickup patterns

#Process the Data:
bison_clean = bison_recipe |> prep() |> juice()

#Define linear model:
model_lm = linear_reg() |> set_engine("lm")

#Define cross validation for experimentation with other regression forms later
bison_cv = bison_train |> vfold_cv(v=5)

#Define Workflow:
fit_lm_cv = workflow() |>
  add_model(model_lm) |>
  add_recipe(bisont_recipe) |>
  fit_resamples(bison_cv)

#Fit Model:
fit_lm_cv |> collect_metrics(summarize = F)

# Plot with custom y-axis labels
ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "loess") +
  scale_y_log10(labels = comma) +  # Use comma format for log scale labels
  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
       x = "Year", 
       y = "Count / Cost",
       color = "License Type") +
  theme_minimal()
```
=======
---
title: "ECNS 560 Project - Bison Population"
author: "Kyle Froisland"
date: "2024-10-16"
output: html_document
---

@@

# Plot with custom y-axis labels
ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "loess") +
  scale_y_log10(labels = comma) +  # Use comma format for log scale labels
  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
       x = "Year", 
       y = "Count / Cost",
       color = "License Type") +
  theme_minimal()
```
>>>>>>> ae5d3f30ef6b3c153d8c3a489f3f9dc3d1e461cb

### 2024-12-02 Draft Exploratory Analysis.Rmd
Add files via upload

---
title: "Draft Exploratory Analysis"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

@@

  In further exploration, other indicators such as weather or local policy differences may be more useful indicators of population levels.

>>>>

<<<<<<< HEAD
---
title: "Draft Exploratory Analysis"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

What we see from the initial visualization is that there seems to be a general downward trend in Bison population from 2002 to 2017, but that there is no obvious trend between sates in between. In other words, while all states ultimately have lower population numbers in 2017 than in 2002, there is fluctuation in between those years in which some states saw relatively steady decline (such as Montana), others saw sharp drops amid periods of steady decline (such as North Dakota), and some even saw short periods of population growth (such as South Dakota and Wyoming).

  After realizing that there is some variety in the details of population growth or decline across states, we sought to confirm our suspicion that the overall population trend is negative. What we see in the following graph is a confirmation of that suspicion, where population sees a strong downward trend across the Great Plains:
![Great Plains Population Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/bison pop over time by state and binned.png)
In this plot, each point on this plot represents a state for a given year, and the color represents the population bin that the state falls into. We can see that the number of blue (high population) and green (medium population) observations decreases over time, as the number of orange (low population) observations increases over time. This shows that states are generally trending down from higher population bins to lower population bins and this general trend is captured by the downward sloping local regression line.

  In further exploration, other indicators such as weather or local policy differences may be more useful indicators of population levels.
=======
---
title: "Draft Exploratory Analysis"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

@@

  In further exploration, other indicators such as weather or local policy differences may be more useful indicators of population levels.
>>>>>>> ae5d3f30ef6b3c153d8c3a489f3f9dc3d1e461cb
